\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology

### 2017-07-26 mers-structure.tex
Minor revisions

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
%\tbc{Fields' Virology textbook defines `terminal host' as `one that is not involved in maintaining the virus, such infections rarely result in transmission to other human or non-human hosts'. I think you want a different word here. Maybe `transient host'?}
%\gdc{Good point about the term. Hope the definition leans towards `one not involved in maintenance' rather than `limited to index case'. Asked a friend working on zoonoses about this.}
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.
% \tbc{Powerful and succinct abstract. Great.}

@@

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines %\lmc{how about a citation here?}
\citep{who_2016_MERS}.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.

@@

The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we %\lmc{estimate?}
estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels %\lmc{is ``allow'' the best word here? I understand you' rather not use ``biases'' or ``misleads'', though...}
the model cannot dismiss humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly %\lmc{maybe ``non-uniformly'' is redundant here} \gdc{disagree, the point I am trying to make is that the data are mostly human and that human data are further skewed in how they are sampled}
sampled human cases and are robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.

@@

\subsection*{Signals of MERS-CoV seasonality}
Unlike the most simpler %\lmc{simpler?}
``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, 
% please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'. 
structured population models possess the ability to estimate migration times. % So do Markov Jumps. 
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

@@

\subsection*{Demographic trajectory of MERS-CoV in camels}
Since MERS-CoV in humans will have a distinct epidemiology
% evolves under a distinct demographic regime, %% evolves under a demographic regime?
we split our sequence dataset based on whether sequences were collected in humans or camels.
We supplement camel sequence data with a single earliest sequence from each human cluster, treating viral diversity present in humans as a sentinel sample of MERS-CoV diversity circulating in camels.
This removes conflicting demographic signals sampled during human outbreaks, where densely sampled closely related sequences from humans could be misconstrued as evidence of demographic crash in the viral population.

@@

\section*{Acknowledgements}
We would like to thank Allison Black for useful discussion and advice.
GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.

>>>>

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.

@@

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines %\lmc{how about a citation here?}
\citep{who_2016_MERS}.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data (174 MERS-CoV genomes from human infections and 100 MERS-CoV genomes from camel infections) to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.

@@

The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we %\lmc{estimate?}
estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot dismiss humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly %\lmc{maybe ``non-uniformly'' is redundant here} \gdc{disagree, the point I am trying to make is that the data are mostly human and that human data are further skewed in how they are sampled}
sampled human cases and are robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
\tbc{I think this sentence (``We therefore...'') needs to be spelled out a bit more. It feels like the logic is what the Monte Carlo sampling does. Either spell out more, or push back to section with figure \ref{ABC}.}

@@

\subsection*{Signals of MERS-CoV seasonality}
Unlike simpler ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
% please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'.
% TB: Agree.
structured population models possess the ability to estimate migration times. % So do Markov Jumps.
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

@@

\subsection*{Demographic trajectory of MERS-CoV in camels}

\tbc{Slightly strange to have this at the level of full results subsection and not have a main text figure.}
Since MERS-CoV in humans will have a distinct epidemiology
% evolves under a distinct demographic regime, %% evolves under a demographic regime?
we split our sequence dataset based on whether sequences were collected in humans or camels.
We supplement camel sequence data with a single earliest sequence from each human cluster, treating viral diversity present in humans as a sentinel sample of MERS-CoV diversity circulating in camels.
This removes conflicting demographic signals sampled during human outbreaks, where densely sampled closely related sequences from humans could be misconstrued as evidence of demographic crash in the viral population.

@@

\section*{Acknowledgements}
We would like to thank \tbc{XXX, YYY, ZZZ, etc...} for kindly sharing sequence data, and Allison Black for useful discussion and advice.
GD is supported by the Mahan postdoctoral fellowship from the Fred Hutchinson Cancer Research Center.
TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774-01.
